master thesis

### 2018-03-23 Paper - Latex/Paper - Skizze - v4.tex
Paper - Results: rewrite Results motivation and the first parts of Metadata analysis

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Metadata analysis}
    In order to generate a realistic simulation, influencing parameters are investigated. \\
    Since the transitions are minimized in an ancestral state reconstruction, this is an important 
      parameter to consider. On the other hand, the completeness of our input data are influencing 
      values. Therefore, two major types are distinguished:
    \begin{enumerate}
      \item Biological parameters (A result of the evolutionary process.):
        \begin{itemize}
          \item transition probabilities
        \end{itemize}
      \item Distribution of the loss of information:
        \begin{itemize}
          \item Loss of topology ($\rightarrow$ mutlifurcations)
          \item Unknown information about states of some leaf nodes
        \end{itemize}
    \end{enumerate}
    The influence of these parameters are tested on our result using our simulation (Section 
      \ref{sec:simulation}).

@@

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Missing information}

@@

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Ancestral state reconstruction methods}
    As presented in the introduction, there are some methods for ancestral state reconstruction. For 
      this purpose, various studied methods and their advantages and disadvantages are compared below. \\

@@

    Parsimony-based methods are used in this work, since they are fully sufficient for the presented 
      use case here. Following the principle of the simpler model first. \\
    Felsenstein \cite{Felsenstein2003} discusses in his book two algorithms that generalize all 
      previous methods (from Camin and Sokal \cite{Camin1965}, \todo{Kluge and Farris} and Farris 
      \cite{Farris1970}): Fitch parsiomony \cite{Fitch1971} and Sankoff parsimony \cite{Sankoff1975}. \\
    \anmerkungstext{Unter Farris war auch noch der Begriff Wagner trees in gebrauch, als 
      Verallgemeinerung der parsimonious trees von Camin und Sokal. (Lydia)} \\
    \todo{Wagner-parsimony \cite{Swofford1987}} \\

@@

  % %---------------------------------------------------------------------------------------------------
  % %---------------------------------------------------------------------------------------------------
  % %---------------------------------------------------------------------------------------------------
  % \section{real data analysis}
  %   \begin{itemize}
  %     \item Import tree
  %     \item Import interactions
  %     \item run castor algorithm / and others?
  %     \item interprete results (cross validation - leave 100 out)
  %   \end{itemize}

@@

%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
\chapter{Results}
  A big point in this chapter is the result of examining the input data. How is the situation? What 
    influence does that have on our actual result? What can we do about it? Our simulation gave us 
    some results to this. \\
  Otherwise, this chapter is mainly about the actual reconstruction of the states. This means, on 
    one hand investigation of origins and losses of the inner nodes and on the other, the prediction 
    of unknown states of leaf nodes. \\
   
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Metadata analysis - Missing information}
    As previously presented, we have two types of missing information: unknown states of leaf nodes 
      and multifurcation. \\

    We examined the ridge of multifurcation of the tree. A complete phylogenetic tree would be 
      binary, which means the number of leaf nodes is closely to the number of internal nodes. But 
      since we only work with a synthesis tree, this tree is multifurcated: we have 241 974 internal 
      nodes and 2 293 463 leaf nodes. \\
    For a first overview we collected for every node its number of children (degree $-1$), and plotted
      this in two histograms, see figure \ref{fig:childrenOfNodes}. \\
    The multifurcation affects only the internal nodes. We collected the number of children $-2$ of 
      every node (a node with two children is binary). That means it discribes the number of nodes which we have lost from the real (binary) 
    phylogenetic tree. \\
    \begin{figure}[h!]
      \centering
      \begin{subfigure}[b]{0.5\textwidth}
        \includegraphics[width=0.9\textwidth]{Figures/multifurc.pdf}
        \caption{Histogram with automatic binwidth. \\ ~ \\ ~}
      \end{subfigure}
      ~~~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
        %(or a blank line to force the subfigure onto a new line)
      \begin{subfigure}[b]{0.42\textwidth}
        \includegraphics[trim = 0mm 0mm 30mm 0mm, clip, width=0.9\textwidth]{Figures/multifurc_small.pdf}
        \caption{Histogram with $binwidth = 1$. \\ light blue: binary; \\ dark blue: multifunction}
      \end{subfigure}
      \caption{Histograms about the multifurcation of the internal nodes of the synthesis tree.}
      \label{fig:childrenOfNodes}
    \end{figure}
    As \colorbox{red}{you} can see, we are very far from a binary tree. \\
    Some subtrees have been examined, these have the following percentages of missing information: See 
      table \ref{table:percentage loss information subtrees}.
    \begin{table}[h!]
      \begin{center}
        \begin{tabular}{ |l||r|r| }
          \hline
          Subtree of & unknown states & multifurcation \\ 
          \hline \hline
          Eukaryota       & 97.34~\%  & 98.16~\% \\
          \hline \hline
          Metazoa         & 96.44~\%  & 87.93~\% \\ \hline
          Fungi           & 98.87~\%  & 96.97~\% \\ \hline
          Chloroplastida  & 99.14~\%  & 89.46~\% \\
          \hline \hline            
          % Arthropoda      & 97.49~\%  & 89.95~\% \\ \hline
          Apicomplexa     & 86.26~\%  & 87.16~\% \\ \hline
          Nematoda        & 89.01~\%  & 88.59~\% \\ \hline
          Chordata        & 88.59~\%  & \cellcolor{green!50}66.49~\% \\ \hline
          Platyhelminthes & \cellcolor{green!50}68.73~\%  & 80.34~\% \\
          \hline \hline            
          Insecta         & 97.11~\%  & 90.78~\% \\
          \hline  
        \end{tabular}
      \end{center}
      \caption{Examination of subtrees regarding missing information. \\ 
        % Taxa of subtrees: Domain, Kingdom, Phylum, Class \\ 
        Unknown states: missing state information of leaf nodes \\ 
        Multifurcation: missing internal nodes}
      \label{table:percentage loss information subtrees} 
    \end{table}

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Data artifacts}
      At this point we also found out that there are some nodes with only one child node (55,700 nodes). \\
      These are both, the most nodes are right in front of a leaf, as well as some nodes are deep in the 
        tree (3,956 with height $>2$). They are probably a result from the fact that taxonomic information 
        has been incorporated into a phylogeny. \\
      Some examples:
      \begin{itemize}
        \item Nephroselmidophyceae: (class) \\
          https://tree.opentreeoflife.org/opentree/argus/ottol@1038762
        \item Phrynocrinidae: (family) \\
          https://tree.opentreeoflife.org/opentree/argus/ottol@3647979
        \item Elaeocarpus sylvestris: \\
          https://tree.opentreeoflife.org/opentree/argus/opentree9.1@ott166969
      \end{itemize}

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Taxa}
      The investigation of the taxonomy revealed that our tree has three kingdoms: Chloroplastida, 
        Metazoa, Fungi, 53 phyla, 195 classes and 924 orders. \\
      Since the analysis of the tree is not part of this work, it should be mentioned here that, 
        according to recent findings, this is not complete and we lack some taxa in every rank. For 
        example, Cavalier-Smith says that one distinguishes between seven and nine kingdoms 
        \cite{CavalierSmith1981}. \\
      In section \pageref{subsec:listPhyla} of the appendix \colorbox{red}{you} can find a list of all phyla.

@@

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Results of simulation / Influence of different parameters}

@@

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Biological view}
      To analyze the results, we have selected some phyla (sub-trees) to evaluate our results 
        selectively from the biological point of view: Chordata, Nematoda, Platyhelminthes and 
        Apicomplexa. \\
      In Table \ref{table:phylum leaf nodes states} we compare the given states with the predicted ones. \\
      There are several factors that play a role here, and in part, crystallize in these examples. \\
      An important factor here is that the credibility of the results with the accuracy of the input 
        data, i. the data of GloBI, stands and falls. Errors of incorrect input data can be amplified by 
        incorrect prediction of unknown species and can be reversed in order to improve the data 
        situation of GloBI. \\
      Since we look at such large trees we can not expect to know all the parasites, so we look at 
        individual positives. Positive in the sense that the majority have the opposite state.

@@

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Origins and Losses}

@@

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Cross evaluation - leave 100 out}
      We ran the castor algorithm 100 times with leaving 100 randomized free-living or parasitic species 
        out of the input data to see how stable our result is. Of these 10,000 nodes, 9,238 were unique. 
        Of that, we predicted 9060 ($\approx 98.17~\%$) correctly and 169 ($\approx 1.82~\%$) wrongly, 
        with duplicate draws always having the same prediction. \\
      What is the best way to model this data? We again tested the influence of the taxa and the depth 
        of leaf nodes and calculated the BICs (Table: \ref{table:BIC cross validation}).

@@

>>>>

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Metadata analysis}
    In order to generate a realistic simulation, influencing parameters are investigated. \\
    Since the transitions are minimized in an ancestral state reconstruction, this is an important 
      parameter to consider. On the other hand, the completeness of our input data are influencing 
      values. Therefore, two major types are distinguished:
    \begin{enumerate}
      \item Biological parameters (A result of the evolutionary process.):
        \begin{itemize}
          \item transition probabilities
        \end{itemize}
      \item Distribution of the loss of information:
        \begin{itemize}
          \item Loss of topology ($\rightarrow$ mutlifurcations)
          \item Unknown information about states of some leaf nodes
        \end{itemize}
    \end{enumerate}
    The influence of these parameters are tested on our result using our simulation (section 
      \ref{sec:simulation}).

@@

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Missing information} \label{sec:methods - missing information}

@@

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Ancestral state reconstruction methods} \label{sec:ancestral state reconstruction}
    As presented in the introduction, there are some methods for ancestral state reconstruction. For 
      this purpose, various studied methods and their advantages and disadvantages are compared below. \\

@@

    Parsimony-based methods are used in this work, since they are fully sufficient for the presented 
      use case here. Following the principle of the simpler model first. \\
    \anmerkungstext{Nochmal darstellen warum und selbst wenn sie fully sufficient wäre, heißt es ja 
      nicht, dass man sie nehmen muss. Also: motiviere! (Renard)} \\
    Felsenstein \cite{Felsenstein2003} discusses in his book two algorithms that generalize all 
      previous methods (from Camin and Sokal \cite{Camin1965}, \todo{Kluge and Farris} and Farris 
      \cite{Farris1970}): Fitch parsiomony \cite{Fitch1971} and Sankoff parsimony \cite{Sankoff1975}. \\
    \anmerkungstext{Unter Farris war auch noch der Begriff Wagner trees in gebrauch, als 
      Verallgemeinerung der parsimonious trees von Camin und Sokal. (Lydia)} \\
    \todo{Wagner-parsimony \cite{Swofford1987}} \\

@@

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{real data analysis}
    \todo{Evaluation section in methods?}
    \begin{itemize}
      \item Import tree
      \item Import interactions
      \item run castor algorithm / and others?
      \item interprete results (cross validation - leave 100 out)
    \end{itemize}
    We then evaluated the results in two ways. On the one hand, we have examined the results of some 
      sub-trees for their credibility with the help of biological background knowledge. On the other 
      hand, we performed a leaf 100 out cross validation.

@@

%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
\chapter{Results}
  This work deals with the ancestral state reconstruction of the entire Eukaryota relatives tree. \\

  For this reconstruction, we first analyzed our data. This is the tree of OTL \cite{Hinchliff2015}
    and the data for the leaf nodes (free-living or parasite states) from GloBI \cite{Poelen2014} 
    (section \ref{sec:results - missing information}). \\
  Next we compared different possible methods. We decided to take a closer look at maximum parsimony 
    algorithms because they are best suited to the problem at hand. The reason for this can be found 
    in section \ref{sec:ancestral state reconstruction}. We tested these on different simulations of 
    the data and compared their predictive power (section \ref{sec:results of simulation}). \\
  We used the Sankoff algorithm \cite{Sankoff1975} implemented by Louca et al. \cite{Louca2017}, 
    which performs best, for the actual reconstruction on the real data. \\

  We then evaluated the results in two ways. On the one hand, we have examined the results of some 
    sub-trees for their credibility with the help of biological background knowledge (Subsection 
    \ref{sec:biological view}). On the other hand, we performed a leaf 100 out cross validation
    (Subsection \ref{sec:results cross evaluation}).

  % The next sections now guide you through the results of these different steps.

%   Otherwise, this chapter is mainly about the actual reconstruction of the states. This means, on 
%     one hand investigation of origins and losses of the inner nodes and on the other, the prediction 
%     of unknown states of leaf nodes. \\
%    
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Metadata analysis - Missing information} \label{sec:results - missing information}
    As previously presented in the methods (section \ref{sec:methods - missing information}), we have 
      two types of missing information: unknown states of leaf nodes and multifurcation. \\

    A tree is multifurcated if there are nodes that have more than two children. In the case of a 
      phylogenetic tree, the ridge of multifurcation describes the amount of lack of information about 
      the topology of the tree. A complete phylogenetic tree would be binary, meaning that the number 
      of leaf nodes equals the number of internal nodes minus one. Since we only work with a 
      artificially constructed tree (synthesis of several phylogenetic trees), this tree is strongly 
      multifurcated: It has 241,974 internal nodes for 2,293,463 leaf nodes $\rightarrow$ 98.16 \% 
      missing nodes. \\
    We calculated these percentages of missing nodes and also missing state information for some 
      subtrees and plotted them in table \ref{table:percentage loss information subtrees}. \\
    \begin{table}[h!]
      \begin{center}
        \begin{tabular}{ |l||r|r| }
          \hline
          Subtree of & unknown states & multifurcation \\ 
          \hline \hline
          Eukaryota       & 97.34~\%  & 98.16~\% \\
          \hline \hline
          Metazoa         & 96.44~\%  & 87.93~\% \\ \hline
          Fungi           & 98.87~\%  & 96.97~\% \\ \hline
          Chloroplastida  & 99.14~\%  & 89.46~\% \\
          \hline \hline            
          % Arthropoda      & 97.49~\%  & 89.95~\% \\ \hline
          Apicomplexa     & 86.26~\%  & 87.16~\% \\ \hline
          Nematoda        & 89.01~\%  & 88.59~\% \\ \hline
          Chordata        & 88.59~\%  & \cellcolor{green!50}66.49~\% \\ \hline
          Platyhelminthes & \cellcolor{green!50}68.73~\%  & 80.34~\% \\
          \hline \hline            
          Insecta         & 97.11~\%  & 90.78~\% \\
          \hline  
        \end{tabular}
      \end{center}
      \caption{Examination of subtrees regarding missing information. \\ 
        The percentage values show the ratio of missing information of: unknown states (missing state 
          information of leaf nodes) and multifurcation (missing internal nodes). \\
        The subtrees are from different taxa: domain (Eukaryota), kingdom (Metazoa, Fungi, 
          Chloroplastida), phylum (Apicomplexa, Nematoda, Chordata, Platyhelminthes) and class 
          (Insecta). \\
        The two by far smallest values were highlighted in green.} 
      \label{table:percentage loss information subtrees} 
    \end{table}

    We also have for a first overview we collected for each node its number of children (degree $-1$), 
      and plotted this in two histograms, see figure \ref{fig:childrenOfNodes}. \\
    \begin{figure}
      \centering
      \begin{subfigure}[b]{0.59\textwidth}
        \includegraphics[width=0.9\textwidth]{Figures/multifurc.pdf}
        \caption{Histogram with automatic binwidth. \\ ~}
      \end{subfigure}
      \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[trim = 0mm 0mm 30mm 0mm, clip, width=0.9\textwidth]{Figures/multifurc_small.pdf}
        \caption{Histogram with $binwidth = 1$. Higher multifurcation than 30 has been cut off. \\ light blue: binary; \\ dark blue: multifurcation}
      \end{subfigure}
      \caption{Histograms about the multifurcation of the internal nodes of the synthesis tree. \\
        For each node, the number of children (degree $-1$) was collected. A node is multifurcated 
        if it has more than two children, so we deducted two from each number of children. We have 
        plotted this in two histograms: number of children $-2$ on the x-axis with log scale and the 
        number of nodes with this amount on the y-axis.}
      \label{fig:childrenOfNodes}
    \end{figure}
    The multifurcation affects only the internal nodes. We collected the number of children $-2$ of 
      every node (a node with two children is binary). That means it discribes the number of nodes 
      which we have lost from the real (binary) phylogenetic tree. \\
    It can be recognized that we are very far from a binary tree. \\

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Data artifacts}
      At this point we also found out that there are nodes with only one child node (55,700 nodes). \\
      The most of these nodes are next to a leaf node, others are deep in the tree (3,956 with height 
        $>2$). They are probably a result from the fact that taxonomic information has been incorporated 
        into a phylogeny. \\
      \anmerkungstext{probably? Da ist eine Spekulation hier? Kann man das nicht herausfinden? 
        Spekulieren solltest du am ehesten erst in der Diskussion. (Thilo)} \\
      Some examples of these nodes are:
      \begin{itemize}
        \item Nephroselmidophyceae: (class) \\
          https://tree.opentreeoflife.org/opentree/argus/ottol@1038762
        \item Phrynocrinidae: (family) \\
          https://tree.opentreeoflife.org/opentree/argus/ottol@3647979
        \item Elaeocarpus sylvestris: \\
          https://tree.opentreeoflife.org/opentree/argus/opentree9.1@ott166969
      \end{itemize}

@@

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Results of simulation / Influence of different parameters} \label{sec:results of simulation}

@@

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Biological view} \label{sec:biological view}
      To analyze the results, we have selected some phyla (sub-trees) to evaluate our results 
        selectively from the biological point of view: Chordata, Nematoda, Platyhelminthes and 
        Apicomplexa. \\
      In Table \ref{table:phylum leaf nodes states} we compare the given states with the predicted ones. \\
      There are several factors that play a role here, and in part, crystallize in these examples. \\
      An important factor here is that the credibility of the results with the accuracy of the input 
        data, i. the data of GloBI, stands and falls. Errors of incorrect input data can be amplified by 
        incorrect prediction of unknown species and can be reversed in order to improve the data 
        situation of GloBI. \\
      Since we look at such large trees we can not expect to know all the parasites, so we look at 
        individual positives. Positive in the sense that the majority have the opposite state.

@@

      %---------------------------------------------------------------------------------------------------
      \subsubsection{Origins and Losses}

@@

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Cross evaluation - leave 100 out} \label{sec:results cross evaluation}
      We ran the castor algorithm 100 times with leaving 100 randomized free-living or parasitic species 
        out of the input data to see how stable our result is. Of these 10,000 nodes, 9,238 were unique. 
        Of that, we predicted 9060 ($\approx 98.17~\%$) correctly and 169 ($\approx 1.82~\%$) wrongly, 
        with duplicate draws always having the same prediction. \\
      What is the best way to model this data? We again tested the influence of the taxa and the depth 
        of leaf nodes and calculated the BICs (Table: \ref{table:BIC cross validation}).

@@

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Data situation}

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Taxa}
      The investigation of the taxonomy revealed that the OTL tree has three kingdoms: Chloroplastida, 
        Metazoa, Fungi, 53 phyla, 195 classes and 924 orders. \\
      Since the analysis of the tree is not part of this work, it should be mentioned here that, 
        according to recent findings, this is not complete and we lack some taxa in every rank. For 
        example, Cavalier-Smith says that one distinguishes between seven and nine kingdoms 
        \cite{CavalierSmith1981}. \\
      In section \pageref{subsec:listPhyla} of the appendix is a list of all phyla.
